Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-06-01 SIMPaper/sbpaper.Rmd
finalized paper

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
revised: 2022-06-01
accepted: 2022-07-01
abstract: "Neural network-based survival methods have been developed to model data-driven covariate interactions. While these methods have led to better predictive performance than regression-based approaches, they cannot model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) as a new approach that combines the case-base sampling framework with flexible architectures. Our method naturally accounts for censoring, does not require additional hyperparameters and uses binary cross-entropy as the loss function. Using a novel sampling scheme and data augmentation, we incorporate time directly into a feed-forward neural network. CBNN predicts the probability of an event occurring at a given moment and estimates the hazard function. We compare the performance of CBNN to survival methods based on regression and neural networks in two simulations and two real data applications. We report two time-dependent metrics for each model. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time and outperforms the competing neural network approaches. For a simple simulation with an exponential hazard model, CBNN consistently outperforms the other neural network methods. For a complex simulation, which highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, CBNN outperforms all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. This study highlights the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of survival outcomes."

@@

# Introduction
Smooth-in-time accelerated failure time (AFT) models can estimate absolute risks by modeling the hazard directly through user-specified baseline hazard distribution [@kleinbaum2012survival]. Cox proportional hazards models are used more often than AFT models, causing analyses to be based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009]. The identification of an appropriate distribution for the baseline hazard in an AFT model may be difficult for common diseases that have many interacting risk factors or a Cox model whose pathogenesis may change with age, making it difficult to maintain the proportional hazards assumption. For example, previous studies of breast cancer incidence have discovered time-varying interactions with covariates of interest, including tumor size [@coradini2000time]. One approach to resolve this involves using the basis of splines on time to provide flexibility in the baseline hazard [@royston2002flexible]. However, regression-based models are limited in that they require prior knowledge about potential time-varying interactions and their quantitative effects.

@@

We note that these alternative neural network approaches all require custom loss functions [@katzman2018DeepSurv] [@lee2018DeepHit] [@dsmPaper]. DeepHit introduces a new hyperparameter weighing its two loss functions (negative log-likelihood and ranking losses) while DSM requires a two-phase learning process and user implementations for distributions beyond Log-Normal or Weibull  [@lee2018DeepHit] [@dsmPaper]. Regression-based approaches require prior specification of all interaction terms, which makes it challenging to model covariate effects that change over time. The current neural network models provide flexibility at the cost of opacity, while regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying interactions and a flexible baseline hazard using commonly available neural network components. Our approach to modeling the full hazard uses case-base sampling [@hanley2009]. This sampling technique allows probabilistic models to predict survival outcomes. As part of the case-base framework, we use transformations of time as a feature (covariate) to specify different baseline hazards. For example, by including splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model [@royston2002flexible], however, this still requires explicit use of time-varying interactions. CBNN can model both without user specification; only a sufficiently deep network is required.

@@

In this section, we define case-base sampling, which converts the total survival time into discrete person-specific moments (person-moments). Then, we detail how neural networks can be used within this framework, explicitly incorporating time as a feature while adjusting for the sampling bias. Finally, we report on the software versions used. A package is available for use at [GITHUBLINK]. The entire code base to generate the analyses in this paper is available at [GITHUB LINK].

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward component, to which an offset term (to adjust for bias from case-base sampling) is added, then passed through a sigmoid activation function (Figure \ref{fig:NNarch}). We use the sigmoid function as its inverse is the odds, which we can use to calculate the hazard. The general form for the neural network using CBNN is:

@@

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By approximating a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility of the model. We provide a detailed description of the choices we made in the next sub-section.

@@

Neural networks are flexible when defining the architecture and optimization parameters. These hyperparameter decisions can affect the estimated parameters and were chosen during a set of initial simulations to determine if CBNN can learn complex interactions in practice. We set a max number of epochs to be $2000$, batch size as $512$, learning rate as $10^{-3}$, decay as $10^{-7}$, patience as $10$ epochs, $\{50,50,25,25\}$ nodes in each hidden layer with $50\%$ dropout at each layer, a validation split of $20\%$, a testing split of $10%$, a minimum delta loss on the validation set of $10^{-7}$ over 10 epochs and Adam [@gulli2017] as our optimizer. These choices may permit the model to approximate higher-order interactions while preventing over-fitting [@srivastava2014dropout]. We fix a train-validation-test split that allows us to update the weights with a subset of the data (training), assess performance at each epoch (validation) and gauge performance of the final model (test) for each method. We select the best weights after training based on validation loss [@gulli2017].

@@

R [@Rsoft] and python [@py] are used to evaluate methods from both languages. We fit the Cox model using the **survival** package [@survpkg], the CBLR model using the **casebase** package [@cbpkg], the DeepSurv model using the **survivalmodels** package [@survmods], the DeepHit model using **pycox** [@lee2018DeepHit] and the DSM model using **DeepSurvivalMachines** [@dsmPaper]. We made the components of CBNN using the **casebase** package for the sampling step and the **keras** [@keras] package for our neural network architecture. The **simsurv** package [@simsurv] is used for our simulation studies, while **flexsurv** [@flexsurv] is used to fit a flexible baseline hazard using splines for our complex simulation. We use the implementation of $C_{IPCW}$??? from the python package **sksurv** [@sksurv]. The **riskRegression** package [@riskRegression] is used to get the Index of Prediction Accuracy (IPA metric). Both metrics are described in detail in the following section. We modify the **riskRegression** package to be used with any user supplied risk function $F$. To ensure that both R and Python-based models are running in unison on the same data through our simulations and bootstrap, we use the **reticulate** package [@reticulate].

@@

where $\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2$ and $\psi$ are basis splines. To generate the $gamma$ coefficients, let $g$: $\mathbb{R} \rightarrow \mathbb{R}$ be a smoothing method for variable $t$ by a projection on to a set of basis functions on splines: 
\srb{i used gt as the function for splines. Specifically, natural cubic spline function of log time https://www.rdocumentation.org/packages/flexsurv/versions/2.1/topics/flexsurvspline . therefore, should i use g(log(t)) and and replace "smoothing method" with natural cubic spline? SAHIR: will fix any issues in this section in the final version.}
\begin{align}
g(t)=\sum^{m}_{l=1}\psi_{l} (t)\gamma_{l}.
\end{align}

An initial model with three knots ($m=5$) is fit using the *flexsurvspline* function with an intercept-only model on the data, which returns the coefficients ($\gamma$) and basis of time ($\psi$) we use in our model. Note that we fix these values for the analysis. The $\beta$ coefficients represent direct effects, $\tau_{2}$ and ${\tau_3}$ represent interactions and $\tau_{1}$ is a time-varying interaction.

>>>>

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
revised: 2022-06-01
accepted: 2022-07-01
abstract: "Neural network-based survival methods have been developed to model data-driven covariate interactions. While these methods have led to better predictive performance than regression-based approaches, they cannot model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) as a new approach that combines the case-base sampling framework with flexible architectures. Our method naturally accounts for censoring, does not require additional hyperparameters and uses binary cross-entropy as the loss function. Using a novel sampling scheme and data augmentation, we incorporate time directly into a feed-forward neural network. CBNN predicts the probability of an event occurring at a given moment and estimates the hazard function. We compare the performance of CBNN to survival methods based on regression and neural networks in two simulations and two real data applications. We report two time-dependent metrics for each model. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time and outperforms the competing neural network approaches. For a simple simulation with an exponential hazard model, CBNN outperforms the other neural network methods. For a complex simulation, which highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, CBNN outperforms all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. This study highlights the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of survival outcomes. An R package is available at [https://github.com/Jesse-Islam/cbnn](https://github.com/Jesse-Islam/cbnn)."

@@

# Introduction
Smooth-in-time accelerated failure time (AFT) models can estimate absolute risks by modeling the hazard directly through a user-specified baseline hazard distribution [@kleinbaum2012survival]. Cox proportional hazards models are used more often than AFT models, causing analyses to be based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009]. The identification of an appropriate distribution for the baseline hazard in an AFT model may be difficult for common diseases that have many interacting risk factors, or a Cox model where the disease pathogenesis may change with age, making it difficult to maintain the proportional hazards assumption. For example, previous studies of breast cancer incidence have discovered time-varying interactions with covariates of interest, such as tumor size [@coradini2000time]. One approach to provide flexibility in the baseline hazard involves using the basis of splines on time in our model [@royston2002flexible]. However, regression-based models are limited in that they require prior knowledge about potential time-varying interactions and their quantitative effects.

@@

We note that these alternative neural network approaches all require custom loss functions [@katzman2018DeepSurv] [@lee2018DeepHit] [@dsmPaper]. DeepHit introduces a hyperparameter weighing its two loss functions (negative log-likelihood and ranking losses) while DSM requires a two-phase learning process and user implementations for distributions beyond Log-Normal or Weibull  [@lee2018DeepHit] [@dsmPaper]. Regression-based approaches require prior specification of all interaction terms, which makes it challenging to model covariate effects that change over time. The current neural network models provide flexibility at the cost of opacity, while regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying interactions and a flexible baseline hazard using commonly available neural network components. Our approach to modeling the full hazard uses case-base sampling [@hanley2009]. This sampling technique allows probabilistic models to predict survival outcomes. As part of the case-base framework, we use transformations of time as a feature (covariate) to specify different baseline hazards. For example, by including splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model [@royston2002flexible], however, this still requires explicit use of time-varying interactions. CBNN can model both without extra tuning parameters.

@@

In this section, we define case-base sampling, which converts the total survival time into discrete person-specific moments (person-moments). Then, we detail how neural networks can be used within this framework, explicitly incorporating time as a feature while adjusting for the sampling bias. Finally, we report on the software versions used. An R package is available for use at [https://github.com/Jesse-Islam/cbnn](https://github.com/Jesse-Islam/cbnn). The entire code base to generate the analyses in this paper is available at [https://github.com/Jesse-Islam/cbnnManuscript](https://github.com/Jesse-Islam/cbnnManuscript).

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward component, to which an offset term is added, then passed through a sigmoid activation function (Figure \ref{fig:NNarch}). We use the sigmoid function as its inverse is the odds, which we can use to calculate the hazard. The general form for the neural network using CBNN is:

@@

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By approximating a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility of the fukk hazard function. We provide a detailed description of the choices we made in the next sub-section.

@@

Neural networks are flexible when defining the architecture and optimization parameters. These hyperparameter decisions can affect the estimated parameters and were chosen during a set of initial simulations to determine if CBNN can learn complex interactions in practice. We set a max number of epochs to be $2000$, batch size as $512$, learning rate as $10^{-3}$, decay as $10^{-7}$, patience as $10$ epochs, $\{50,50,25,25\}$ nodes in each hidden layer with $50\%$ dropout at each layer, a minimum delta loss on the validation set of $10^{-7}$ over 10 epochs and Adam [@gulli2017] as our optimizer. These choices may permit the model to approximate higher-order interactions while preventing over-fitting [@srivastava2014dropout]. We fix a train-validation-test split that allows us to update the weights with a subset of the data (training), assess performance at each epoch (validation) and gauge performance of the final model (test) for each method. We select the best weights after training based on validation loss [@gulli2017].

@@

R [@Rsoft] and python [@py] are used to evaluate methods from both languages. We fit the Cox model using the **survival** package [@survpkg], the CBLR model using the **casebase** package [@cbpkg], the DeepSurv model using the **survivalmodels** package [@survmods], the DeepHit model using **pycox** [@lee2018DeepHit] and the DSM model using **DeepSurvivalMachines** [@dsmPaper]. We made the components of CBNN using the **casebase** package for the sampling step and the **keras** [@keras] package for our neural network architecture. The **simsurv** package [@simsurv] is used for our simulation studies, while **flexsurv** [@flexsurv] is used to fit a flexible baseline hazard using splines for our complex simulation. We use the implementation of $C_{IPCW}$ from the python package **sksurv** [@sksurv]. The **riskRegression** package [@riskRegression] is used to get the Index of Prediction Accuracy (IPA metric). Both metrics are described in detail in the following section. We modify the **riskRegression** package to be used with any user supplied risk function $F$. To ensure that both R and Python-based models are running in unison on the same data through our simulations and bootstrap, we use the **reticulate** package [@reticulate].

@@

where $\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2$ and $\psi$ are basis splines. To generate the $gamma$ coefficients, an initial cubic splines model with three knots is fit using the *flexsurvspline* function with an intercept-only model on the data, which returns the coefficients ($\gamma$) we use in our model. Note that we fix these values for the analysis. The $\beta$ coefficients represent direct effects, $\tau_{2}$ and ${\tau_3}$ represent interactions and $\tau_{1}$ is a time-varying interaction.

### 2022-06-01 SIMPaper/sbpaper.tex
finalized paper

\abstract{Neural network-based survival methods have been developed to
model data-driven covariate interactions. While this flexibility has led
to methods with better predictive performance than regression-based
approaches, current neural networks cannot model both time-varying
interactions and complex baseline hazards. To address this, we propose
Case-Base Neural Networks (CBNN) as a new approach that combines the
case-base sampling framework with flexible architectures. Our method
naturally accounts for censoring, does not require additional
hyperparameters and uses binary cross-entropy as the loss function.
Using a novel sampling scheme and data augmentation, we incorporate time
directly into a feed-forward neural network. CBNN predicts the
probability of an event occurring at a given moment and estimates the
hazard function. We compare the performance of CBNN to survival methods
based on regression and neural networks in two simulations and two real
data applications. We report two time-dependent metrics for each model.
In the simulations and real data applications, CBNN provides a more
consistent predictive performance across time and outperforms the
competing neural network approaches. For a simple simulation with an
exponential hazard model, CBNN consistently outperforms the other neural
network methods. For a complex simulation, which highlights the ability
of CBNN to model both a complex baseline hazard and time-varying
interactions, CBNN outperforms all competitors. The first real data
application shows CBNN outperforming all neural network competitors,
while a second real data application shows competitive performance. This
study highlights the benefit of combining case-base sampling with deep
learning to provide a simple and flexible modeling framework for
data-driven, time-varying interaction modeling of survival outcomes.}

@@

Smooth-in-time accelerated failure time (AFT) models can estimate
absolute risks by modeling the hazard directly through user-specified
baseline hazard distribution \citep{kleinbaum2012survival}. Cox
proportional hazards models are used more often than AFT models, causing
analyses to be based on hazard ratios and relative risks rather than on
survival curves and absolute risks \citep{hanley2009}. The
identification of an appropriate distribution for the baseline hazard in
an AFT model may be difficult for common diseases that have many
interacting risk factors or a Cox model whose pathogenesis may change
with age, making it difficult to maintain the proportional hazards
assumption. For example, previous studies of breast cancer incidence
have discovered time-varying interactions with covariates of interest,
including tumor size \citep{coradini2000time}. One approach to resolve
this involves using the basis of splines on time to provide flexibility
in the baseline hazard \citep{royston2002flexible}. However,
regression-based models are limited in that they require prior knowledge
about potential time-varying interactions and their quantitative
effects.

@@

We note that these alternative neural network approaches all require
custom loss functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}
\citep{dsmPaper}. DeepHit introduces a new hyperparameter weighing its
two loss functions (negative log-likelihood and ranking losses) while
DSM requires a two-phase learning process and user implementations for
distributions beyond Log-Normal or Weibull \citep{lee2018DeepHit}
\citep{dsmPaper}. Regression-based approaches require prior
specification of all interaction terms, which makes it challenging to
model covariate effects that change over time. The current neural
network models provide flexibility at the cost of opacity, while
regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method
that models time-varying interactions and a flexible baseline hazard
using commonly available neural network components. Our approach to
modeling the full hazard uses case-base sampling \citep{hanley2009}.
This sampling technique allows probabilistic models to predict survival
outcomes. As part of the case-base framework, we use transformations of
time as a feature (covariate) to specify different baseline hazards. For
example, by including splines of time as covariates, we can approximate
the Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}, however, this still requires explicit use
of time-varying interactions. CBNN can model both without user
specification; only a sufficiently deep network is required.

@@

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Then, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Finally, we report on the software versions used. A
package is available for use at {[}GITHUBLINK{]}. The entire code base
to generate the analyses in this paper is available at {[}GITHUB
LINK{]}.

@@

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the model. We provide a detailed description
of the choices we made in the next sub-section.

@@

Neural networks are flexible when defining the architecture and
optimization parameters. These hyperparameter decisions can affect the
estimated parameters and were chosen during a set of initial simulations
to determine if CBNN can learn complex interactions in practice. We set
a max number of epochs to be \(2000\), batch size as \(512\), learning
rate as \(10^{-3}\), decay as \(10^{-7}\), patience as \(10\) epochs,
\(\{50,50,25,25\}\) nodes in each hidden layer with \(50\%\) dropout at
each layer, a validation split of \(20\%\), a testing split of \(10%
\), a minimum delta loss on the validation set of \(10^{-7}\) over 10
epochs and Adam \citep{gulli2017} as our optimizer. These choices may
permit the model to approximate higher-order interactions while
preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test) for each
method. We select the best weights after training based on validation
loss \citep{gulli2017}.

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DeepHit model using \textbf{pycox}
\citep{lee2018DeepHit} and the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper}. We made the components
of CBNN using the \textbf{casebase} package for the sampling step and
the \textbf{keras} \citep{keras} package for our neural network
architecture. The \textbf{simsurv} package \citep{simsurv} is used for
our simulation studies, while \textbf{flexsurv} \citep{flexsurv} is used
to fit a flexible baseline hazard using splines for our complex
simulation. We use the implementation of \(C_{IPCW}\)??? from the python
package \textbf{sksurv} \citep{sksurv}. The \textbf{riskRegression}
package \citep{riskRegression} is used to get the Index of Prediction
Accuracy (IPA metric). Both metrics are described in detail in the
following section. We modify the \textbf{riskRegression} package to be
used with any user supplied risk function \(F\). To ensure that both R
and Python-based models are running in unison on the same data through
our simulations and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

where
\(\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2\)
and \(\psi\) are basis splines. To generate the \(gamma\) coefficients,
let \(g\): \(\mathbb{R} \rightarrow \mathbb{R}\) be a smoothing method
for variable \(t\) by a projection on to a set of basis functions on
splines:
\srb{i used gt as the function for splines. Specifically, natural cubic spline function of log time https://www.rdocumentation.org/packages/flexsurv/versions/2.1/topics/flexsurvspline . therefore, should i use g(log(t)) and and replace "smoothing method" with natural cubic spline? SAHIR: will fix any issues in this section in the final version.}
\begin{align}
g(t)=\sum^{m}_{l=1}\psi_{l} (t)\gamma_{l}.
\end{align}

An initial model with three knots (\(m=5\)) is fit using the
\emph{flexsurvspline} function with an intercept-only model on the data,
which returns the coefficients (\(\gamma\)) and basis of time (\(\psi\))
we use in our model. Note that we fix these values for the analysis. The
\(\beta\) coefficients represent direct effects, \(\tau_{2}\) and
\({\tau_3}\) represent interactions and \(\tau_{1}\) is a time-varying
interaction.

>>>>

\abstract{Neural network-based survival methods have been developed to
model data-driven covariate interactions. While these methods have led
to better predictive performance than regression-based approaches, they
cannot model both time-varying interactions and complex baseline
hazards. To address this, we propose Case-Base Neural Networks (CBNN) as
a new approach that combines the case-base sampling framework with
flexible architectures. Our method naturally accounts for censoring,
does not require additional hyperparameters and uses binary
cross-entropy as the loss function. Using a novel sampling scheme and
data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in two simulations and two real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a simple simulation with an exponential hazard model,
CBNN outperforms the other neural network methods. For a complex
simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions, CBNN outperforms all
competitors. The first real data application shows CBNN outperforming
all neural network competitors, while a second real data application
shows competitive performance. This study highlights the benefit of
combining case-base sampling with deep learning to provide a simple and
flexible modeling framework for data-driven, time-varying interaction
modeling of survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

Smooth-in-time accelerated failure time (AFT) models can estimate
absolute risks by modeling the hazard directly through a user-specified
baseline hazard distribution \citep{kleinbaum2012survival}. Cox
proportional hazards models are used more often than AFT models, causing
analyses to be based on hazard ratios and relative risks rather than on
survival curves and absolute risks \citep{hanley2009}. The
identification of an appropriate distribution for the baseline hazard in
an AFT model may be difficult for common diseases that have many
interacting risk factors, or a Cox model where the disease pathogenesis
may change with age, making it difficult to maintain the proportional
hazards assumption. For example, previous studies of breast cancer
incidence have discovered time-varying interactions with covariates of
interest, such as tumor size \citep{coradini2000time}. One approach to
provide flexibility in the baseline hazard involves using the basis of
splines on time in our model \citep{royston2002flexible}. However,
regression-based models are limited in that they require prior knowledge
about potential time-varying interactions and their quantitative
effects.

@@

We note that these alternative neural network approaches all require
custom loss functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}
\citep{dsmPaper}. DeepHit introduces a hyperparameter weighing its two
loss functions (negative log-likelihood and ranking losses) while DSM
requires a two-phase learning process and user implementations for
distributions beyond Log-Normal or Weibull \citep{lee2018DeepHit}
\citep{dsmPaper}. Regression-based approaches require prior
specification of all interaction terms, which makes it challenging to
model covariate effects that change over time. The current neural
network models provide flexibility at the cost of opacity, while
regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method
that models time-varying interactions and a flexible baseline hazard
using commonly available neural network components. Our approach to
modeling the full hazard uses case-base sampling \citep{hanley2009}.
This sampling technique allows probabilistic models to predict survival
outcomes. As part of the case-base framework, we use transformations of
time as a feature (covariate) to specify different baseline hazards. For
example, by including splines of time as covariates, we can approximate
the Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}, however, this still requires explicit use
of time-varying interactions. CBNN can model both without extra tuning
parameters.

@@

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Then, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Finally, we report on the software versions used. An R
package is available for use at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to
generate the analyses in this paper is available at
\url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the fukk hazard function. We provide a
detailed description of the choices we made in the next sub-section.

@@

Neural networks are flexible when defining the architecture and
optimization parameters. These hyperparameter decisions can affect the
estimated parameters and were chosen during a set of initial simulations
to determine if CBNN can learn complex interactions in practice. We set
a max number of epochs to be \(2000\), batch size as \(512\), learning
rate as \(10^{-3}\), decay as \(10^{-7}\), patience as \(10\) epochs,
\(\{50,50,25,25\}\) nodes in each hidden layer with \(50\%\) dropout at
each layer, a minimum delta loss on the validation set of \(10^{-7}\)
over 10 epochs and Adam \citep{gulli2017} as our optimizer. These
choices may permit the model to approximate higher-order interactions
while preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test) for each
method. We select the best weights after training based on validation
loss \citep{gulli2017}.

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DeepHit model using \textbf{pycox}
\citep{lee2018DeepHit} and the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper}. We made the components
of CBNN using the \textbf{casebase} package for the sampling step and
the \textbf{keras} \citep{keras} package for our neural network
architecture. The \textbf{simsurv} package \citep{simsurv} is used for
our simulation studies, while \textbf{flexsurv} \citep{flexsurv} is used
to fit a flexible baseline hazard using splines for our complex
simulation. We use the implementation of \(C_{IPCW}\) from the python
package \textbf{sksurv} \citep{sksurv}. The \textbf{riskRegression}
package \citep{riskRegression} is used to get the Index of Prediction
Accuracy (IPA metric). Both metrics are described in detail in the
following section. We modify the \textbf{riskRegression} package to be
used with any user supplied risk function \(F\). To ensure that both R
and Python-based models are running in unison on the same data through
our simulations and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

where
\(\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2\)
and \(\psi\) are basis splines. To generate the \(gamma\) coefficients,
an initial cubic splines model with three knots is fit using the
\emph{flexsurvspline} function with an intercept-only model on the data,
which returns the coefficients (\(\gamma\)) we use in our model. Note
that we fix these values for the analysis. The \(\beta\) coefficients
represent direct effects, \(\tau_{2}\) and \({\tau_3}\) represent
interactions and \(\tau_{1}\) is a time-varying interaction.
